The Design and Implementation of the Synchronous Language \CEU

### 2016-04-02 paper.tex
...

\begin{abstract}
\CEU is a synchronous language inspired by Esterel which aims to offer a 
simpler semantics and more fine-grained control over the program execution.
%
\CEU employs an event-triggered notion of time that allows for a simple
reasoning about reactions in parallel and enables compile-time analysis to 
produce deterministic and concurrency-safe programs.
%
We discuss the particularities of our design in comparison to Esterel, such as 
stack-based internal events, temporal analysis for concurrent statements, safe 
integration with \C, and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our clean semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

\CEU is a synchronous reactive language inspired by Esterel in which programs 
advance in a sequence of discrete reactions to external events.
%
%Regarding the similarities with Esterel,
%\CEU has a strong imperative flavor, with explicit control flow through 
%sequences, loops, and parallel compositions.
%
Like Esterel, \CEU is designed for control-intensive applications, supporting 
concurrent lines of execution, known as \emph{trails}, and broadcast 
communication through events.
%
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
An \code{await} is the only statement that halts a running reaction and allows 
a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be directly implemented.

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel \ax and 
\CEU \bx for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 5--10, in both 
implementations), i.e., Esterel's `$\|$' and \CEU's \code{par/and} constructs 
are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 4--11), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 4--13):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events, depending on external stimuli the environment provides.
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The event-triggered execution of a program in \CEU is as 
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an occurring event completely before handling the next.
%
Based on the synchronous hypothesis, a program takes no time on step 2 and is 
always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running (step 2), it is enqueued to occur in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in Esterel and 
\CEU.
The box \code{Real World} shows event occurrences over a continuous timeline 
divided in units of 10 milliseconds.
Esterel and \CEU use discrete logical units of time in which the occurring 
events (which are the same in all boxes) have to fit somehow:
%
\begin{itemize}
\item \code{[Box-1]}:
``Esterel with fixed-length ticks'',
within which reactions to occurring inputs have to fit~\cite{esterel.fixed}.
%
We assume a \code{R(boot)} reaction at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur ``simultaneously'' during the idle \code{tick-3}.
However, the program can only detect one occurrence of \code{E}, which is 
considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
``Esterel with variable-length ticks'', which are adjusted to the CPU time the 
corresponding reaction takes to complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also results in
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
``\CEU with unique and queue-based input events''.
We also assume a \code{R(boot)} reaction before any input.
%
Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
behavior in \CEU is similar to \code{Box-2} for the first two units of time 
(\code{boot-0} and \code{A-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
\end{itemize}

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), the 
events \code{B} and \code{C} are simultaneous, even though \code{A} and 
\code{B} actually happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example that detects simultaneous button clicks.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events never overlap because they are atomic.
Automated mutual exclusion simplifies the reasoning about the program and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
To cope with deterministic intra-reaction communication, \CEU supports internal 
events which programs can \code{emit} and \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), broadcasts \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
      \emph{(No other trails awake to \code{b}.)}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

Internal events provide a fine-grained intra-reaction control mechanism.
For instance, it brings a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence the calling trail resumes and passes the assertion 
test (ln. 11).

@@

The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
prevents infinite execution by design.
Both sides of the \code{par/or} have an \code{await} statement to avoid 
instantaneous termination (ln. 4,7).
However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
same reaction that reaches them, the \code{par/or} would terminate and restart 
the \code{loop} instantaneously, resulting in infinite execution.

\begin{figure}
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
For instance, sometimes we need to execute a block of code immediately, and 
then periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
%
On the one hand, we transfer the burden of dealing with these specific corner 
cases to the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

From our experience, however, we find that it is not uncommon to require 
stateful and non-commutative calls to execute concurrently.
%
This is the case for logging (e.g., calls to \code{\_printf} in parallel) or 
redrawing objects in the screen.
%
Figure~\ref{fig.redraw}.a shows an abstract code to animate and redraw objects 
\code{back} and \code{fore} in the screen in trails in parallel.
In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
conflicting pixels.
This way, \code{\_redraw} is non commutative and the code does not compile 
correctly.
%
However, since the \code{back} object should be redrawn before the \code{fore} 
object, we want to rely on \CEU's intra-reaction deterministic execution based 
on lexical order.
Therefore, in Figure~\ref{fig.redraw}, we redefine the redrawing function to an 
``intimidating'' name \code{\_redraw\_non\_commutative} to be explicit about 
its effect (ln. 2), and annotate it as safe (ln. 4--5) to compile successfully.

@@

.
\end{lstlisting}
\centering\small{\ax \code{\_redraw} \textbf{cannot} be concurrent}
\end{minipage}
%
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define redraw_non_commutative redraw
end
@safe _redraw_non_commutative with
      _redraw_non_commutative;
par do
    <...> // animate and redraw "back"
        _redraw_non_commutative(back);
with
    <...> // animate and redraw "fore"
        _redraw_non_commutative(fore);
end
\end{lstlisting}
\centering\small{\bx \code{\_redraw\_non\_commutative} \textbf{can} be concurrent}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Making the non-commutative redrawing calls from \ax to compile in \bx.
\label{lst.c}
}
\end{figure}

@@

The design decisions XXX, raise some important limitations as follows:
%
\begin{itemize}
\item Purely reactive execution.
    hardware
A limitation of event-triggered execution is that all program behavior must be purely
reactive, given that no code can execute in the absence of inputs. Tick-triggered execu-
tion allows for active behavior, since code can execute regulary on every tick. Although
it supports asynchronous active execution [Sant’Anna et al. 2012], we still consider
 ́ as purely reactive since its synchronous core cannot express active behavior.
\item bounded-only reactions,
\item sequential
\end{itemize}
%
We believe that a simple-tractable synchronous core
and extend it with memory-isolated parallelizabe asynchronous execution.

    - permeates all aspects
    - As far as we know, \CEU is the first Esterel-based language to HAVE?
      queue-based/unique in the semantics of the language and not in the
      library

@@

In the sections that follow, we discuss the implementation highlights related 
to the peculiarities of \CEU:
the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

@@

\subsection{Single-Threaded Dispatching}

@@

TODO: comparison with Esterel

>>>>

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with support for a simpler 
semantics and more fine-grained execution control.
%
\CEU employs an event-triggered notion of time which enables compile-time 
analysis for simultaneous reactions to produce deterministic and 
concurrency-safe programs.
%
We discuss the particularities of our design in comparison to Esterel, such as 
stack-based internal events, temporal analysis for concurrent statements, safe 
integration with \C, and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
% TODO: simple simple
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

\CEU is a synchronous reactive language inspired by Esterel in which programs 
advance in a sequence of discrete reactions to external events.
%
%Regarding the similarities with Esterel,
%\CEU has a strong imperative flavor, with explicit control flow through 
%sequences, loops, and parallel compositions.
%
Like Esterel, \CEU is designed for control-intensive applications, supporting 
concurrent lines of execution, known as \emph{trails}, and broadcast 
communication through events.
%
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
An \code{await} is the only statement that halts a running reaction and allows 
a program to advance in this discrete notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be directly implemented.

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel \ax and 
\CEU \bx for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 5--10, in both 
implementations), as Esterel's `$\|$' and \CEU's \code{par/and} constructs are 
equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 4--11), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 4--13):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events, depending on external stimuli the environment provides.
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The event-triggered execution of a program in \CEU is as 
follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an occurring event completely before handling the next.
%
Based on the synchronous hypothesis, a program takes a negligible time on step 
2 and is always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running, it is enqueued to occur in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} shows event occurrences over a continuous timeline 
divided in units of 10 milliseconds.
The other boxes define distinct discrete logical units of time in which the 
same occurring events fit differently.
%
\begin{itemize}
\item \code{[Box-1]}:
``Esterel with fixed-length ticks'',
within which reactions must fit~\cite{esterel.fixed}.
%
We assume a \code{R(boot)} reaction at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} is considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
``Esterel with variable-length ticks'', which are adjusted to the CPU time the 
corresponding reaction takes to complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also results in
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
``\CEU with unique and queue-based input events''.
We also assume a \code{R(boot)} reaction before any input.
%
Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
behavior in \CEU is similar to \code{Box-2} for the first two units of time 
(\code{boot-0} and \code{A-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
\end{itemize}

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in Figure~\ref{fig.ticks} (\code{box-1} and \code{box-2}), the 
events \code{B} and \code{C} are considered simultaneous, even though \code{A} 
and \code{B} ``physically'' happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example to detect simultaneous button clicks in 
\CEU.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events never overlap because they are atomic.
Automated mutual exclusion simplifies the reasoning about the program and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

A limitation of event-triggered execution is that all program behavior must be 
purely reactive, given that no code can execute in the absence of inputs.
%
Tick-triggered execution allows for active behavior, since code can execute 
regulary on every tick.
%
Although \CEU supports asynchronous active execution~\cite{ceu.tr}, we still 
consider it as purely reactive since its synchronous core cannot express active 
behavior.

@@

Esterel makes no semantic distinctions between internal and external signals.
%In particular, programs can emit external and internal signals simultaneously, 
%with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
%In \CEU, subsequent external events coming from the environment define a 
%timeline, which programs cannot interfere by emitting new external events.
%
%To cope with deterministic intra-reaction communication, \CEU supports 
%internal events which programs can \code{emit} and \code{await}.
In \CEU, in contrast with queue-based external events, internal events follow a 
stack-based execution policy similar to subroutine calls in typical programming 
languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
(ln. 6,9) in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes, broadcasts \code{b}, and pauses (ln. 4--5).
\item 2nd trail awakes, calls \code{\_g()}, and terminates (ln. 8--9).
      \emph{(No other trails awake to \code{b}.)}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates (ln. 5--6).
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

Internal events define fine-grained micro reactions which stack one on top of 
another inside the reaction to an external event.
%Internal events provide a fine-grained intra-reaction control mechanism.
For instance, it brings a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence the calling trail resumes and passes the assertion 
test (ln. 11).

@@

The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
prevents infinite execution by design.
Both sides of the \code{par/or} have an \code{await} statement to avoid 
instantaneous termination (ln. 4,7) of the enclosing \code{loop} (ln 2--9).
However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
same reaction that reaches them, the \code{par/or} would terminate and restart 
the \code{loop} instantaneously, resulting in infinite execution.
%
\begin{comment}
\begin{figure}
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}
\end{comment}
%
In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
\begin{comment}
For instance, sometimes we need to execute a block of code immediately, and 
then periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
\end{comment}
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

From our experience, however, we find that it is not uncommon to require 
stateful and non-commutative calls to execute concurrently.
%
This is the case for logging (e.g., calls to \code{\_printf} in parallel) or 
redrawing objects in the screen.
%
Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw objects 
\code{back} and \code{fore} in the screen in trails in parallel.
In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
conflicting pixels.
This way, \code{\_redraw} is non commutative and the code does not compile 
correctly.
%
However, since the \code{back} object should be redrawn before the \code{fore} 
object, we want to rely on \CEU's intra-reaction deterministic execution based 
on lexical order.
Therefore, in Figure~\ref{lst.redraw}, we redefine the redrawing function to an 
``intimidating'' name \code{\_redraw\_non\_commutative} to be explicit about 
its effect (ln. 2), and annotate it as safe (ln. 4--5) to compile successfully.

@@

.
\end{lstlisting}
\centering\small{\ax \code{\_redraw} \textbf{cannot} be concurrent}
\end{minipage}
%
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define redraw_non_commutative redraw
end
@safe _redraw_non_commutative with
      _redraw_non_commutative;
par do
    <...> // animate and redraw "back"
        _redraw_non_commutative(back);
with
    <...> // animate and redraw "fore"
        _redraw_non_commutative(fore);
end
\end{lstlisting}
\centering\small{\bx \code{\_redraw\_non\_commutative} \textbf{can} be concurrent}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Making the non-commutative redrawing calls from \ax to compile in \bx.
\label{lst.redraw}
}
\end{figure}

@@

Our synchronous and deterministic approach also leads to some limitations as 
follows:
%
\begin{itemize}
\item Execution is purely reactive as a result of event-triggered reactions.
Since only occurrences of events can start reactions, programs cannot execute 
proactively in the absence of events.
\item Reactions must execute in bounded time due to the synchronous hypothesis.
As a synchronous language, \CEU requires that CPU times for reactions are 
negligible in comparison to the rate of incoming events.
\item Execution is sequential because of intra-reaction determinism.
The deterministic semantics of \CEU does not make parallelization easy (to be 
discussed in Section~\ref{sec.impl.single}).
\end{itemize}
%
Nonetheless, we advocate for a tractable synchronous reactive core that 
supports shared memory with deterministic execution.
%
To deal with the limitations, we recommend for memory-isolated parallelizable 
asynchronous primitives as language extensions, but which are not in the scope 
of this paper.

@@

In the sections that follow, we discuss the implementation highlights related 
to the peculiarities of \CEU:
the temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
single-threaded dispatching (Section~\ref{sec.impl.single}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

@@

\subsection{Single-Threaded Dispatching}
\label{sec.impl.single}

@@

TODO: most about comparison with Esterel
    - TODO: related work?
    - As far as we know, \CEU is the first Esterel-based language to HAVE?
      queue-based/unique in the semantics of the language and not in the
      library
